Pattern matching for\\ $(231,213)$-avoiding permutations

### 2015-02-01 article.tex
modif : utilisation des environnement lemma, proposition, coroolary

		\textbf{Lemma  \numl{firstelement} \thenum} If a permutation is in $\Av{n}{213,231}$ 
		then the first element is either
		the minimal or the maximal element.\\

@@

		\textbf{Corollary  \num \thenum} A permutation $\pi \in \Perm{n}$ is in $\Av{n}{213,231}$ if and only if for $1 \leq i \leq n$, 
		$\pi[i:]$ start either with the maximal element
		or the minimal element of $\pi[i:]$.\\

		\textbf{Corollary \numl{stripe} \thenum}  If $\pi \in \Av{n}{213,231}$ its stripe can be define such as, \\
		$\stripe{\pi}{i} = \left\{ 
  		\begin{array}{l l}
   			\ustep & \quad \text{if $\pi(i)$ is the minimal element of $\pi[i:]$}\\    
   			\dstep & \quad \text{if $\pi(i)$ is the maximal element of $\pi[i:]$}\\
	 	\end{array} \right. 
	 	$\\

		\textbf{Lemma \numl{updown}  \thenum} Let $\pi \in \Av{n}{213,231}$,
		$\forall i,j \in [n]$, 
		$i<j$, 
		if $\stripe{\pi}{i}=\ustep$
		and $\stripe{\pi}{j}=\dstep$
		then $\pi(i)<\pi(j)$.\\

@@

		\textbf{Proof} By induction on the size of the permutation :
		the proposition is true for the permutations $(1,2)$ and $(2,1)$.
		Suppose the proposition true for $\Av{n}{213,231}$.
		Let $\pi \in \Av{n}{213,231}$, 
		we show that the proposition holds for $\pi$.
		Let $\pi'$ be the permutation $\pi[2:]$.
		$\pi' \in \Av{n}{213,231}$, so
		every element of $p'$ verify our lemma (by induction).
		We still need to compare the first element with the rest.
		But the first element
		is the maximal (minimal) element of all the permutation 
		(by Lemma \ref{firstelement}).
		Moreover if the first element is the maximal (minimal)
		it will mapped to an descent (ascent).
		So the proposition is also true
		for the first element.
		$\Box$\\

		\textbf{Lemma \numl{upup} \thenum} Let $\pi \in \Av{n}{213,231}$,
		$\forall i,j \in [n]$, 
		$i<j$, 
		if $\stripe{\pi}{i}=\ustep$
		and $\stripe{\pi}{j}=\ustep$
		then $\pi(i)<\pi(j)$.\\

@@

		\textbf{Lemma \numl{downdown} \thenum}  Let $\pi \in \Av{n}{213,231}$,
		$\forall i,j \in [n]$,
		$i<j$,
		if $\stripe{\pi}{i}=\dstep$
		and $\stripe{\pi}{j}=\dstep$
		then $\pi(i)>\pi(j)$.\\

@@

		\textbf{Proposition \numl{patternMatchingStripe} \thenum} Let $\ptext, \pmotif \in \Av{n}{213,231}$.
		$\pmotif$ is a motif of $\ptext$ if and only if the stripe of $\pmotif$ is a motif of the stripe of $\ptext$.\\

		\textbf{Proof} \\
		$\Rightarrow$ Suppose that $\pmotif$ is a motif of $\ptext$
		by subsequence $s$.
		By definition $s$ normalized is equal to
		$\pmotif$ so they have the same stripe.\\ 
		$\Leftarrow$ 
		Let $\ptext$ be a permutation avoiding $(213,231)$, 
		$S_1$ be its stripe, 
		$\pmotif$  be a permutation avoiding $(213,231)$, 
		and $S_2$ its stripe.
		Let $\phi$ be the function
		that match $S_2$ to $S_1$.
		We define $\varphi$ as, 
		$\forall x \in [|\pmotif|-1]$,
		$\varphi(x) = \phi(x)$,
		and $\varphi(|\pmotif|) = \phi(|\pmotif|-1)+1$
		Suppose that $\exists x,y \in [|\pmotif|]$, 
		$\pmotif(x)<\pmotif(y)$ 
		and $\ptext(\varphi(x))>\ptext(\varphi(y))$,
		without lost of generality we can suppose that $x<y$,
		Let see all the case possible that can make this 
		hypothesis possible :
		\begin{itemize}		
			\item if $\pmotif(x)<\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\ustep$ 
			and $\stripe{\pmotif}{y}=\ustep$
			(if x and y are increasing in $\pmotif$),
			then $\stripe{\ptext}{\varphi(x)} =\ustep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
			(the match are also increasing).
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(x) < \varphi(y)$.
			And thus, by \textbf{Lemma \ref{upup}},
			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)<\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\ustep$ 
			and $\stripe{\pmotif}{y}=\dstep$
			then $\stripe{\ptext}{\varphi(x)}=\ustep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(x) < \varphi(y)$.
			And thus, by \textbf{Lemma \ref{updown}},
			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)>\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\dstep$ 
			and $\stripe{\pmotif}{y}=\dstep$
			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(y) < \varphi(x)$.
			And thus, by \textbf{Lemma \ref{downdown}},
			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)>\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\dstep$ 
			and $\stripe{\pmotif}{y}=\ustep$
			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(y) < \varphi(x)$.
			And thus, by \textbf{Lemma \ref{updown}},
			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
			Which contradict our hypothesis,
			so this case is not possible.$\Box$

@@

		\textbf{Proof}	
		To solve the permutation pattern for a permutation in $\Av{n}{213,231}$, 
		we find a find a match of the stripe of the pattern in the stripe of the text.
		This can be done linearly by a greedy algorithm, 
		by matching a step whenever it is possible.
		And thank to the definition of the definition of a stripe in \textbf{Corollary \ref{stripe}},
		we do not have to compute the stripe advance.
		Therefore we have an linear (online) algorithm.$\Box$\\

@@

			\textbf{Proposition  \num \thenum} Let $\pmotif$, $\ptext$ 
			be two permutation,
			with  $\pmotif$ a permutation avoiding (231,213), 
			we can decide in $O(|\pmotif|*|\ptext|^2)$ time
			and $O(|\pmotif|*|\ptext|^2)$ space
			if $\pmotif$ 
			is a motif of $\ptext$.\\

@@

			\textbf{Proposition \num \thenum } Let $\bmotif$ be a bivincular motif 
			avoiding (231,213), 
			$\ptext$ a permutation,
			we can decide in $O(|\pmotif|*|\ptext|^3)$ time
			and $O(|\ptext|^3)$ space
			if $\pmotif$ 
			appear in $\ptext$.\\

@@

	\textbf{Proposition  \numl{longestIncreasingSubsequence} \thenum}
	If $s$ is a longest subsequence avoiding (213,231) with last element at index $f$ in $\pi$ then
	$P(\pi)$ is a longest increasing subsequence with last element at index $f$ and
	$M(\pi)$ is a longest decreasing subsequence with last element at index $f$.\\

@@

	\textbf{Proposition  \num \thenum}
	Given a permutation $\pi$, 
	finding the longest subsequence avoiding (231,213)
	can be done in $O(|\ptext|*log(log(|\ptext|)))$ time and in $O(n)$ space.\\

	\textbf{Proof} The proposition \ref{longestIncreasingSubsequence} lead to algorithm 
	where we have to compute longest increasing and decreasing subsequence ending at every index. Then finding the maximum sum of longest increasing and decreasing subsequence ending at the same index.
	Computing the longest increasing and decreasing can be done in $O(|\ptext|*log(log(|\ptext|)))$ time and $O(n)$ space (see \cite{Bespamyatnikh00enumeratinglongest}), then finding the maximum can be done in linear time.$\Box$\\

@@

	\textbf{Proposition \num \thenum } Given two permutation $\pi_1$ and $\pi_2$
	The longest common subsequence
	avoiding (231,213) can be solve in $O(|\pi_1|^3*|\pi_2|^3)$.\\

>>>>

		\begin{lemma}
		\label{lemma:firstelement}
		If a permutation is in $\Av{n}{213,231}$ 
		then the first element is either
		the minimal or the maximal element.
		\end{lemma}

		%\textbf{Lemma  \numl{firstelement} \thenum}

@@

		\begin{corollary}
		\label{corollary:stripe}
		A permutation $\pi \in \Perm{n}$ is in $\Av{n}{213,231}$ if and only if for $1 \leq i \leq n$, 
		$\pi[i:]$ start either with the maximal element
		or the minimal element of $\pi[i:]$.
		\end{corollary}

		\begin{corollary}
		If $\pi \in \Av{n}{213,231}$ its stripe can be define such as, \\
		$\stripe{\pi}{i} = \left\{ 
  		\begin{array}{l l}
   			\ustep & \quad \text{if $\pi(i)$ is the minimal element of $\pi[i:]$}\\    
   			\dstep & \quad \text{if $\pi(i)$ is the maximal element of $\pi[i:]$}\\
	 	\end{array} \right. 
	 	$
		\end{corollary}

		\begin{lemma}
		\label{lemma:updown}
		Let $\pi \in \Av{n}{213,231}$,
		$\forall i,j \in [n]$, 
		$i<j$, 
		if $\stripe{\pi}{i}=\ustep$
		and $\stripe{\pi}{j}=\dstep$
		then $\pi(i)<\pi(j)$.
		\end{lemma}

@@

		\textbf{Proof} By induction on the size of the permutation :
		the proposition is true for the permutations $(1,2)$ and $(2,1)$.
		Suppose the proposition true for $\Av{n}{213,231}$.
		Let $\pi \in \Av{n}{213,231}$, 
		we show that the proposition holds for $\pi$.
		Let $\pi'$ be the permutation $\pi[2:]$.
		$\pi' \in \Av{n}{213,231}$, so
		every element of $p'$ verify our lemma (by induction).
		We still need to compare the first element with the rest.
		But the first element
		is the maximal (minimal) element of all the permutation 
		(by Lemma \ref{lemma:firstelement}).
		Moreover if the first element is the maximal (minimal)
		it will mapped to an descent (ascent).
		So the proposition is also true
		for the first element.
		$\Box$\\

		\begin{lemma}
		\label{lemma:upup}
		Let $\pi \in \Av{n}{213,231}$,
		$\forall i,j \in [n]$, 
		$i<j$, 
		if $\stripe{\pi}{i}=\ustep$
		and $\stripe{\pi}{j}=\ustep$
		then $\pi(i)<\pi(j)$.
		\end{lemma}

@@

		\begin{lemma}
		\label{lemma:downdown}
 		Let $\pi \in \Av{n}{213,231}$,
		$\forall i,j \in [n]$,
		$i<j$,
		if $\stripe{\pi}{i}=\dstep$
		and $\stripe{\pi}{j}=\dstep$
		then $\pi(i)>\pi(j)$.
		\end{lemma}

@@

		\begin{proposition}
		\label{proposition:patternMatchingStripe}
		Let $\ptext, \pmotif \in \Av{n}{213,231}$.
		$\pmotif$ is a motif of $\ptext$ if and only if the stripe of $\pmotif$ is a motif of the stripe of $\ptext$.
		\end{proposition}

		\textbf{Proof} \\
		$\Rightarrow$ Suppose that $\pmotif$ is a motif of $\ptext$
		by subsequence $s$.
		By definition $s$ normalized is equal to
		$\pmotif$ so they have the same stripe.\\ 
		$\Leftarrow$ 
		Let $\ptext$ be a permutation avoiding $(213,231)$, 
		$S_1$ be its stripe, 
		$\pmotif$  be a permutation avoiding $(213,231)$, 
		and $S_2$ its stripe.
		Let $\phi$ be the function
		that match $S_2$ to $S_1$.
		We define $\varphi$ as, 
		$\forall x \in [|\pmotif|-1]$,
		$\varphi(x) = \phi(x)$,
		and $\varphi(|\pmotif|) = \phi(|\pmotif|-1)+1$
		Suppose that $\exists x,y \in [|\pmotif|]$, 
		$\pmotif(x)<\pmotif(y)$ 
		and $\ptext(\varphi(x))>\ptext(\varphi(y))$,
		without lost of generality we can suppose that $x<y$,
		Let see all the case possible that can make this 
		hypothesis possible :
		\begin{itemize}		
			\item if $\pmotif(x)<\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\ustep$ 
			and $\stripe{\pmotif}{y}=\ustep$
			(if x and y are increasing in $\pmotif$),
			then $\stripe{\ptext}{\varphi(x)} =\ustep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
			(the match are also increasing).
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(x) < \varphi(y)$.
			And thus, by \textbf{Lemma \ref{lemma:upup}},
			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)<\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\ustep$ 
			and $\stripe{\pmotif}{y}=\dstep$
			then $\stripe{\ptext}{\varphi(x)}=\ustep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(x) < \varphi(y)$.
			And thus, by \textbf{Lemma \ref{lemma:updown}},
			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)>\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\dstep$ 
			and $\stripe{\pmotif}{y}=\dstep$
			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(y) < \varphi(x)$.
			And thus, by \textbf{Lemma \ref{lemma:downdown}},
			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)>\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\dstep$ 
			and $\stripe{\pmotif}{y}=\ustep$
			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(y) < \varphi(x)$.
			And thus, by \textbf{Lemma \ref{lemma:updown}},
			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
			Which contradict our hypothesis,
			so this case is not possible.$\Box$

@@

		\textbf{Proof}	
		To solve the permutation pattern for a permutation in $\Av{n}{213,231}$, 
		we find a find a match of the stripe of the pattern in the stripe of the text.
		This can be done linearly by a greedy algorithm, 
		by matching a step whenever it is possible.
		And thank to the definition of the definition of a stripe in \textbf{Corollary \ref{corollary:stripe}},
		we do not have to compute the stripe advance.
		Therefore we have an linear (online) algorithm.$\Box$\\

@@

			\begin{proposition}
			Let $\pmotif$, $\ptext$ 
			be two permutation,
			with  $\pmotif$ a permutation avoiding (231,213), 
			we can decide in $O(|\pmotif|*|\ptext|^2)$ time
			and $O(|\pmotif|*|\ptext|^2)$ space
			if $\pmotif$ 
			is a motif of $\ptext$.
			\end{proposition}

@@

			\begin{proposition}
			Let $\bmotif$ be a bivincular motif 
			avoiding (231,213), 
			$\ptext$ a permutation,
			we can decide in $O(|\pmotif|*|\ptext|^3)$ time
			and $O(|\ptext|^3)$ space
			if $\pmotif$ 
			appear in $\ptext$.
			\end{proposition}

@@

	\begin{proposition}
	\label{proposition:longestIncreasingSubsequence}
	If $s$ is a longest subsequence avoiding (213,231) with last element at index $f$ in $\pi$ then
	$P(\pi)$ is a longest increasing subsequence with last element at index $f$ and
	$M(\pi)$ is a longest decreasing subsequence with last element at index $f$.	
	\end{proposition}

@@

	\begin{proposition}
	Given a permutation $\pi$, 
	finding the longest subsequence avoiding (231,213)
	can be done in $O(|\ptext|*log(log(|\ptext|)))$ time and in $O(n)$ space.	
	\end{proposition}

	\textbf{Proof} The proposition \ref{proposition:longestIncreasingSubsequence} lead to algorithm 
	where we have to compute longest increasing and decreasing subsequence ending at every index. Then finding the maximum sum of longest increasing and decreasing subsequence ending at the same index.
	Computing the longest increasing and decreasing can be done in $O(|\ptext|*log(log(|\ptext|)))$ time and $O(n)$ space (see \cite{Bespamyatnikh00enumeratinglongest}), then finding the maximum can be done in linear time.$\Box$\\

@@

	\begin{proposition}
	Given two permutation $\pi_1$ and $\pi_2$
	The longest common subsequence
	avoiding (231,213) can be solve in $O(|\pi_1|^3*|\pi_2|^3)$.	
	\end{proposition}
